Reliable Web Services Composition: A MDD Approach

### 2013-09-30 SOAP_Track2013/src/paperSOAP_tosubmit.tex
inicial --martin

%% This is file `elsarticle-template-1a-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1a-num.tex 151 2009-10-08 05:18:25Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1a-num.tex $
%%
\documentclass[preprint,12pt]{elsarticle}
\usepackage{geometry}
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\journal{Journal of Systems and Software}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{$\pi$SOD-M: A Methodology for Building Service-Oriented Applications in
the Presence of Non-Functional Properties}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author[inst1]{Pl\'acido A. Souza Neto}
\author[inst2]{Genoveva Vargas-Solar}
\author[inst3]{Martin A. Musicante}
\author[inst4]{Valeria~de~Castro}
\author[inst3]{Umberto Costa}

\address[inst1]{Federal Institute of Rio Grande do Norte -- Natal-RN,
Brazil}

\address[inst2]{Universit\'e de Grenoble -- Saint Martin d'H\`{e}res, France}

\address[inst3]{Federal University of Rio Grande do Norte -- Natal-RN,
Brazil}

\address[inst4]{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain}

\begin{abstract}
This paper presents\ldots
\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% main text

@@

Service oriented development methodologies providing models, best practices, and reference architectures to build services' based applications mainly address  functional aspects \cite{1,2,decastro1,PapazoglouH06}.  Non-functional aspects concerning services' and application's "semantics", often expressed as requirements and constraints in general purpose methodologies, are not fully considered or they are added once the application has been implemented in order to ensure some level of reliability (e.g., data privacy, exception handling, atomicity, data persistence). This leads to services' based applications that are partially specified and that are thereby partially compliant with application requirements.

The objective of this work   is to model non-functional constraints and associate them to  services' based applications  early during the services' composition modeling phase. Therefore this paper presents $\pi$-SOD-M, a model-driven method  that extends the SOD-M  \cite{decastro1} for building reliable  services' based information systems (SIS). 
%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.

Our work proposes to extend the SOD-M \cite{decastro1} method with  (i)  the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing non-functional constraints associated to services' based applications.  
%{\em A-policies} are used to express constraints which can be applied to all the services' composition  or to a particular service used for implementing it. They represent both systems' cross-cutting aspects (e.g., exception handling expressing what to do when a service is not available) and use constraints imposed by the services  (e.g., the fact that a service requires imposes an authentication protocol for executing a method). 
Our work  also (ii) defines the $\pi$-{\sc Pews}  meta-model \cite{Placido2010LTPD} providing guidelines for expressing the composition and the {\em A-policies}. Finally, our work (iii) defines model to model transformation rules for generating the $\pi$-{\sc Pews} model of a reliable services' composition starting from the extended services' composition model; and, model to text transformations for generating the corresponding implementation. As will be shown within our environment implementing these meta models and rules, one may represent both systems' cross-cutting aspects (e.g., exception handling for describing what to do when a service is not available, recovery, persistence aspects) and constraints associated to services, that must be respected for using them (e.g., the fact that a service requires an authentication protocol for executing a method).

The remainder of the paper is organized as follows. Section \ref{sec:motivation} gives an overview of our approach. It describes a motivation example that integrates and synchronizes well-known social networks services namely Facebook, Twitter and, Spotify. Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and \ref{sec:mmrules} describe respectively the three key elements of our proposal, namelly the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation rules that support the semi-automatic generation of reliable services' compositions.
%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
%
Section \ref{sec:implementation} describes implementation and validation issues.
Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms. Section \ref{sec:conclusions} concludes the paper and discusses future work.

%*********************************************************************************************************
\section{Modeling reliable services' compositions with $\pi$-SOD-M}\label{sec:motivation}
%*********************************************************************************************************
Consider for instance the following scenario. An organization wants to provide the services' based application "To Publish Music" that monitors the music a person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts. Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify.
For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
\begin{itemize}
\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export a method for  updating the status of a given user:
\begin{itemize} 
%\item{\sf\small get-Status ( usedid ): String ; 
\item {\sf\small update-Status ( userid, new-status ): String}; 
\end{itemize}
\end{itemize}
%
%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
%the scenario. The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed. 
%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).

Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
The "To Publish Music" scenario 
%(showed by doted line) 
starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\begin{figure}
\centering
\includegraphics[width=0.55\textwidth]{figs/SC}
\caption{BPMN model of the "To Publish Music" scenario}
\label{fig:E3valuemodel}
\end{figure}

@@

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Modeling a services' based application}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach   providing  a set of guidelines to build services' based information systems (SIS) \cite{decastro1,decastro2}.  Therefore, SOD-M proposes to use services as first-class objects for the whole process of the SIS  development and it  follows a Model Driven Architecture (MDA) \cite{miller}  approach. Extending from the highest level of abstraction of the MDA, SOD-M provides  a conceptual structure to: first, capture the system requirements and specification in high-level abstraction models (computation independent models, CIM’s); next,  starting from such models build platform independent models (PIM’s) specifying the system details; next transform such models into platform specific models (PSM’s) that bundles the specification of the system with the details of the targeted platform; and finally, serialize such model into the working-code that implements the system. 
\begin{figure} [htpb]
\centering
\includegraphics[width=0.65\textwidth]{figs/SODM}
\caption{SOD-M development process}
\label{fig:sodm}
\end{figure} 
As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by building the high-level computational independent models and enables specific models for a service platform to be obtained as a result  \cite{decastro1}. Referring to the "To Publish Music" application, using SOD-M the designer starts defining an E3value model \footnote{The E3 value model is a business model that represents a business case %graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) 
and allows  to understand the environment in which the services' composition will be placed \cite{e3value}.}  at the CIM level and then the corresponding models of the PIM are generated leading to a services' composition model (SCM).
%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
%provide
%is an evolving approach to software development that deals with the provision of 
%models, transformations between them and code generators to address software development. 
%One of the main advantages of model-driven approaches is the provision of 
%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.

Now, consider that besides the services' composition that represents the order in which the services are called for implementing the application "To Publish Music" it is necessary to model  other requirements that represent the (i) conditions imposed by services for being contacted, for example the fact the Facebook and Twitter require authentication protocol in order to call their methods for updating the wall; (ii) the conditions stemming from the business rules of the application logic, for example the fact that the walls in Facebook and Twitter must show the same song title and if this is not possible then none of them is updated.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Modeling non-functional constraints of services' based applications}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
Adding non-functional requirements and services constraints in the services' composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.

@@

%
%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints. 
%
Our work extends SOD-M for building applications by modeling the application logic and its associated non-functional constraints and thereby ensuring the generation of reliable services' composition. 
%This method is described in the following sections.
As a first step in our approach, and for the sake of simplicity we started modeling non-functional constraints at the PSM level. Thus, in this paper we  propose the $\pi$-SCM, the services' composition meta-model extended with {\em A-policies} for modeling non-functional constraints (highlighted in Figure  \ref{fig:sodm} and described in Section \ref{sec:piscm}).  $\pi$-SOD-M defines the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}). Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).

%*********************************************************************************************************
\section{$\pi$ services' composition meta-model}\label{sec:piscm}
%*********************************************************************************************************

@@

 \begin{figure}[htpb]
\centering
\includegraphics[width=0.80\textwidth]{figs/E-service-composition-metamodel}
\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
\label{fig:e-scomposition-metamodel}
\end{figure}

To illustrate the use of the $\pi$-SCM meta-model we used it for defining the {\em A-policy} based composition model of the "To Publish Music" scenario (see
Figure \ref{fig:servicecompositionmodel}). There are three external business collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em italics} to refer to concrete values of the classes of a model that are derived from the classes of a meta-model.}). It also shows the business process  of the  "To Publish Music" application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}. Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.

@@

\begin{figure}[htpb]
\centering
\includegraphics[width=0.70\textwidth]{figs/e-composition-model}
\caption{Services' composition model for the business service "To publish music"}
\label{fig:servicecompositionmodel}
\end{figure}

@@

Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code. For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.

@@

Figure \ref{fig:transformations} shows the transformation principle between the elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model. There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.

@@

%*********************************************************************************************************
\section{Implementation issues}\label{sec:implementation}
%*********************************************************************************************************
This section  describes the $\pi$-SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions. For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application. Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$-SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.

>>>>

\documentclass{sig-alternate}
  \pdfpagewidth=8.5truein
  \pdfpageheight=11truein

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{SAC'13}{March 18-22, 2013, Coimbra, Portugal.}
\CopyrightYear{2013} % Allows default copyright year (2002) to be over-ridden - IF NEED BE.
\crdata{978-1-4503-1656-9/13/03}  % Allows default copyright data (X-XXXXX-XX-X/XX/XX) to be over-ridden.
% --- End of Author Metadata ---

\title{Reliable Web Services Composition: A MDD Approach}

% \subtitle{[Extended Abstract]
% \titlenote{A full version of this paper is available as
% \textit{Author's Guide to Preparing ACM SIG Proceedings Using
% \LaTeX$2_\epsilon$\ and BibTeX} at
% \texttt{www.acm.org/eaddress.htm}}}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{1} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
% 
% 1st. author
% \alignauthor
% Pl\'acido A. Souza Neto\\
% 	   \affaddr{Federal Institute of Rio Grande do Norte (IFRN)},\\
%        \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
%        \affaddr{Natal-RN, Brazil}\\
%        \email{placido.neto@ifrn.edu.br}
% \alignauthor Genoveva Vargar-Solar\\
%        \affaddr{French Council of Scientific Research}\\
%        \affaddr{LIG-LAFMIA }\\
%        \affaddr{Saint Martin d'H\`{e}res, France}\\
%        \email{Genoveva.Vargas-Solar@imag.fr}
% \alignauthor Valeria de Castro\\
%        \affaddr{Universidad Rey Juan Carlos}\\
%        \affaddr{M\'{o}stoles, Spain}\\ 
%        \email{Valeria.deCastro@urjc.es}
% \and
% \alignauthor Martin A. Musicante\\
%        \affaddr{Federal University of Rio Grande do Norte (UFRN)}\\
%        \affaddr{Natal-RN, Brazil}\\
%        \email{mam@dimap.ufrn.br}
% \alignauthor Jos\'e-Luis Zechinelli-Martini\\
%        \affaddr{Fundaci\'on Universidad de las Am\'ericas Puebla (UDLAP)}\\
%        \affaddr{San Andr\'es Cholula, M\'exico}\\
%        \email{joseluis.zechinelli@udlap.mx}
% \alignauthor Christine Collet\\
%        \affaddr{Grenoble Institute of Technology}\\
%        \affaddr{Saint Martin d'H\`{e}res, France}\\
%        \email{Christine.Collet@grenoble-inp.fr}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.

% \additionalauthors{Additional authors: Javier A. Espinosa-Oviedo (Grenoble
% Institute of Technology, email: {\texttt{Javier.Espinosa@imag.fr}}) and
% Esperanza Marcos (Universidad Rey Juan Carlos, email: {\texttt{esperanza.marcos@urjc.es}}).}
 \date{21 September 2012}

% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
This paper presents an approach for modeling and associating {\em Policies} to 
services' based applications. It proposes to extend the SOD-M model driven
method with (i) the $\pi$-{\sc SCM}, a {\em Policy} services' composition
meta-model for representing non-functional constraints associated to services'
based applications; (ii) the $\pi$-{\sc Pews} meta-model providing guidelines
for expressing the composition and the policies; and, (iii)  model to model and
model to text transformation rules for semi-automatizing the implementation of
reliable services' compositions. As will be shown within our environment
implementing these meta models and rules, one may represent both systems'
cross-cutting aspects (e.g., exception handling for describing what to do when a
service is not available, recovery, persistence aspects) and constraints
associated to services, that must be respected for using them (e.g., the fact
that a service requires an authentication protocol for executing a method).
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]

%\terms{Delphi theory}

\keywords{Methodology, $\pi$SOD-M, Sevice Composition, Policy}

%*********************************************************************************************************
\section{Introduction}
%*********************************************************************************************************

@@

Service oriented development methodologies providing mod\-els, best practices,
and reference architectures to build services' based applications mainly address 
functional aspects \cite{1,2,decastro1,PapazoglouH06}.  Non-functional aspects
concerning services' and application's "semantics", often expressed as
requirements and constraints in general purpose methodologies, are not fully
considered or they are added once the application has been implemented in order 
to ensure some level of reliability (e.g., data privacy, exception handling,
atomicity, data persistence). This leads to services' based applications that
are partially specified and that are thereby partially compliant with
application requirements.

The objective of this work is to model non-functional constraints and associate
them to  services' based applications  early during the services' composition
modeling phase. Therefore this paper presents $\pi$SOD-M, a model-driven
meth\-od that extends the SOD-M  \cite{decastro1} for building reliable  services' based
information systems (SIS).
%SOD-M defines a process  starting with the  identification of business services through business modeling, and, by means of models' transformations it allows to obtain a services' composition model \cite{decastro1} and the executable code that implements it.

This work (i) proposes to extend the SOD-M \cite{decastro1} method with 
the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing
non-func\-tion\-al constraints associated to services' based applications; (ii)
defines the $\pi$-{\sc Pews}  meta-model providing
guidelines for expressing the composition and the {\em A-policies}; and finally,
 (iii) defines model to model transformation rules for generating the $\pi$-{\sc
 Pews} model of a reliable services' composition starting from the extended
 services' composition model; and, model to text transformations for generating
 the corresponding implementation. As will be shown within our environment
 implementing these meta models and rules, one may represent both systems'
 cross-cutting aspects (e.g., exception handling for describing what to do when
 a service is not available, recovery, persistence aspects) and constraints
 associated to services, that must be respected for using them (e.g., the fact
 that a service requires an authentication protocol for executing a method).

The remainder of the paper is organized as follows. Section \ref{sec:motivation}
gives an overview of our approach. It describes a motivation example that
integrates and synchronizes well-known social networks services namely Facebook,
Twitter and, Spotify. Sections \ref{sec:piscm}, \ref{sec:pewsmetamodel}, and
\ref{sec:mmrules} describe respectively the three key elements of our proposal,
namely the $\pi$-SCM and $\pi$-{\sc Pews} meta-models and the transformation
rules that support the semi-automatic generation of reliable services'
compositions.
%describes $\pi$-SOD-M method that enables the representation and association of {\em A-policies} to services' composition  thereby making them reliable. 
%
Section \ref{sec:implementation} describes implementation and validation issues.
Section \ref{sec:related} analyses related work concerning policy/contract based programming and, services' composition platforms. Section \ref{sec:conclusions} concludes the paper and discusses future work.

%*********************************************************************************************************
\section{Modeling reliable services' \\ com\-po\-si\-tions with
$\pi$SOD-M}\label{sec:motivation}
%*********************************************************************************************************
Consider for instance the following scenario. An organization wants to provide
the services' based application ``To Publish Music'' that monitors the music a
person is listening during some periods of time and sends the song title  to this person's Twitter and Facebook accounts. Thus, this social network user will have her status synchronized in  Twitter and Facebook (i.e., either the same title is published in both accounts or it is not updated) with the title of the music she is listening in Spotify. For developing this services' based application it is necessary to compose the following services  calling  their  exported methods:
\begin{itemize}
\item The music service   Spotify exports a method for obtaining information  about the music a given user is listening:
\begin{itemize} \item {\sf\small get-Last-Song ( userid ): String} ; \end{itemize}
\item Facebook and Twitter services export a method for  updating the status of a given user:
\begin{itemize} 
%\item{\sf\small get-Status ( usedid ): String ; 
\item {\sf\small update-Status ( userid, new-status ): String}; 
\end{itemize}
\end{itemize}
%
%Figure \ref{fig:E3valuemodel} shows the E3 value model \cite{e3value} of
%the scenario. The E3 value model is a business model that represents a business case graphically as a set of value exchanges ($\nabla$$\triangle$) and value activities (rounded boxes) performed by business actors (squared boxes) and allows us to understand the environment in which the services' composition will be placed. 
%The model in Figure \ref{fig:E3valuemodel} shows Spotify and a private application (which is also a service) that directly interact with users for providing free services for listening and publishing information about music being listened by users. The private application interacts with Spotify for obtaining free information about the flow of music being listened by a user in return of a fee (i.e., premium subscription). Finally, the private application interacts with Facebook and Twitter for updating the user's status and thereby they share non material benefits (i.e., the fact that users subscribe to their networks and are active on them thanks to the private application).

Figure \ref{fig:E3valuemodel} shows the BPMN model  \footnote{Details on BPMN (Business Process Management Notation) can be found in http://www.bpmn.org/} of the scenario. 
The ``To Publish Music'' scenario 
starts by contacting the music service Spotify for retrieving the user's  musical status (activity {\sf Get Song}). 
Twitter and Facebook services are then contacted in parallel for updating the user's status with the corresponding song title (activities {\sf Update Twitter} and {\sf Update Facebook}).
\begin{figure} 
\centering  
\includegraphics[width=0.32\textwidth]{figs/SC} 
\caption{BPMN model of the ``To Publish Music'' scenario}
\label{fig:E3valuemodel}   
\end{figure}

@@

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Modeling a services' based application}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
 Figure \ref{fig:sodm} shows  SOD-M that defines a service oriented approach  
 providing  a set of guidelines to build services' based information systems
 (SIS) \cite{decastro1,decastro2}.  Therefore, SOD-M proposes to use services as
 first-class objects for the whole process of the SIS  development and it 
 follows a Model Driven Architecture (MDA) \cite{miller}  approach. Extending
 from the highest level of abstraction of the MDA, SOD-M provides  a conceptual
 structure to: first, capture the system requirements and specification in
 high-level abstraction models (computation independent models, CIM's); next, 
 starting from such models build platform independent models (PIM's) specifying
 the system details; next transform such models into platform specific models
 (PSM's) that bundles the specification of the system with the details of the
 targeted platform; and finally, serialize such model into the working-code that
 implements the system.

\begin{figure} [htpb]
\centering
\includegraphics[width=0.5\textwidth]{figs/SODM}
\caption{SOD-M development process}
\label{fig:sodm}
\end{figure}

As shown in Figure \ref{fig:sodm}, the SOD-M model-driven process begins by
building the high-level computational independent models and enables specific
models for a service platform to be obtained as a result  \cite{decastro1}.
Referring to the ``To Publish Music'' application, using SOD-M the designer
starts defining an E3value model \footnote{The E3 value model is a business model that
represents a business case and allows  to understand the environment in which
the services' composition will be placed \cite{e3value}.}  at the CIM level and
then the corresponding models of the PIM are generated leading to a services'
composition model (SCM).

%SOD-M proposes a set of models, :  i) the three different MDA abstraction levels: CIM, PIM and PSM; and ii) SOD-M views: business and information system views. 
%Model-Driven Engineering (MDE) \cite{schmidt} and particularly MDA \footnote{Model Driven Architecture (MDA)  is the particular model-driven proposal defined by the Object Management Group (OMG).} 
%provide
%is an evolving approach to software development that deals with the provision of 
%models, transformations between them and code generators to address software development. 
%One of the main advantages of model-driven approaches is the provision of 
%It also provides a conceptual structure where the models used by business managers and analysts can be traced towards more detailed models used by software developers.

Now, consider that besides the services' composition that represents the order
in which the services are called for implementing the application ``To Publish
Music'' it is necessary to model  other requirements that represent the (i)
conditions imposed by services for being contacted, for example the fact the
Facebook and Twitter require authentication protocol in order to call their
methods for updating the wall; (ii) the conditions stemming from the business
rules of the application logic, for example the fact that the walls in Facebook
and Twitter must show the same song title and if this is not possible then none
of them is updated.

%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
\subsection{Modeling non-functional constraints of services' based applications}
%..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--..--
Adding non-functional requirements and services con\-straints in the services'
composition is a complex task that implies programming  protocols for instance authentication protocols to call Facebook and Twitter in our example, and atomicity (exception handling and recovery) for ensuring a true synchronization of the song title disseminated in the walls of the user's Facebook and Twitter accounts.

@@

%
%Our proposal, called $\pi$-SOD-M, extends the services' composition model of the SOD-M method  with the notion of {\em A-Policy} \cite{Espinosa-Oviedo2011a} for representing services' composition constraints. 
%
This work extends SOD-M for building applications by modeling the application
logic and its associated non-func\-tion\-al constraints and thereby ensuring the
generation of reliable services' composition.
%This method is described in the following sections.
As a first step in our approach, and for the sake of simplicity we started
modeling non-functional constraints at the PSM level. Thus, in this paper we
propose the $\pi$-SCM, the services' composition meta-model extended with {\em
A-policies} for modeling non-functional constraints (highlighted in Figure 
\ref{fig:sodm} and described in Section \ref{sec:piscm}).  $\pi$SOD-M defines
the $\pi$-{\sc Pews}  meta-model providing guidelines for expressing the
services' composition and the {\em A-policies} (see Section \ref{sec:pewsmetamodel}), and also defines model to model transformation rules for generating  $\pi$-{\sc Pews} models starting from $\pi$-SCM models that will support executable code generation (see Section \ref{sec:mmrules}). Finally, our work defines model to text transformation rules for generating the program that implements both the services' composition and the associated {\em A-policies} and that is executed by an adapted engine (see Section \ref{sec:implementation}).

%*********************************************************************************************************
\section{$\pi$-services' composition \\ meta-\-model}\label{sec:piscm}
%*********************************************************************************************************

@@

 \begin{figure*}[htpb] 
\centering
\includegraphics[width=0.80	\textwidth]{figs/E-service-composition-metamodel}
\caption{{\em A-policy} based services' composition meta-model ($\pi$-SCM)}
\label{fig:e-scomposition-metamodel}
\end{figure*}

To illustrate the use of the $\pi$-SCM meta-model we used it for defining the
{\em A-policy} based composition model of the ``To Publish Music'' scenario (see
Figure \ref{fig:servicecompositionmodel}). There are three external business
collaborators ({\em Spotify, Twitter} and {\em Facebook} \footnote{We use {\em
italics} to refer to concrete values of the classes of a model that are derived
from the classes of a meta-model.}). It also shows the business process  of the 
``To Publish Music'' application that consists of three service activities: {\em Listen Music}, {\em Public Music} and {\em Confirmation}. Note that  the action {\em Publish Music} of the application calls the actions of two service collaborators namely {\em Facebook} and {\em Twitter}.

@@

@@

\begin{figure}[ht!]
\centering
\includegraphics[width=0.50\textwidth]{figs/e-composition-model}
\caption{Services' composition model for the business service ``To publish
music''}
\label{fig:servicecompositionmodel}
\end{figure}
Thanks to rules and policies  it is possible to model and associate non-functional properties to services' compositions  and then generate the code. For example, the atomic integration of information retrieved from different social network services, automatic generation of an integrated view of the operations executed in different social networks or for providing security in the communication channel when the payment service is called.

@@

Table \ref{tab:transformation} shows the transformation principle between the
elements of the $\pi$-SCM meta-model used for representing the services' composition into the elements of the $\pi$-{\sc Pews} meta-model. There are two groups of rules: those that transform services' composition elements of the $\pi$-SCM to $\pi$-{\sc Pews} meta-models elements; and those that transform rules grouped by policies into {\em A-policy} types.

@@

%*********************************************************************************************************
\section{Implementation issues}\label{sec:implementation}
%*********************************************************************************************************
This section  describes the $\pi$SOD-M development environment that implements the generation of {\em A-policies}' based services' compositions. For a given services' based application, the process  consists in generating the  code starting from a $\pi$-SCM modeling an application. Note that the services' composition model is not modeled from scratch, but it is the result of a general process defined by the $\pi$SOD-M method in which a set of models are built following a service oriented approach \cite{decastro1}.
